Limited predictability of amino acid substitutions in seasonal influenza viruses

### 2020-12-16 article.tex
More details/reference to new SM section

We focus on frequency trajectories that are starting at a zero (low) frequency, \emph{i.e.} $f(t=0)=0$.
These represent new amino acid variants which were absent in the population at the time bin when the trajectory started and are currently rising in the population (see Methods).
Such distinction in novel and ancestral variants is necessary to meaningfully interrogate predictability.
Each rising trajectory of a new mutation implies the existence of another decreasing one at the same position, since frequencies of all amino acids at a given position must sum to one.
%If novel variants arise by selection, we expect to see a stronger signal of selection after conditioning on these novel variants.
In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e., the rapid rise and fix.
The sweep of a mutation can be due to its own fitness effect, to the genetic background or to the effect of the seven other segments.
By considering the ensemble of novel variants that are rising in frequency, we effectively average over backgrounds, obtaining a set of mutations that we expect to be beneficial on average.
If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency should thus enrich for mutations that are positively selected and on their way to fixation.
Interference between multiple such advantages mutations will perturb the trajectories \citep{illingworth_components_2012} but we still expect a average \pierre{replaced ``consistent'' by ``average''} tendency to rise in frequency (see below). \pierre{Is this ``see below'' necessary? I think that's what the reader naturally does! }

@@

However, future distributions for $\Delta t >0$ do not seem to follow a pattern compatible with selective sweeps.
The thick black line in Figure~\ref{fig:trajectories}B shows the average frequency of all trajectories.
This average makes a sharp turn at $t=0$ and is essentially flat for $t>0$ in the case of A/H3N2, and slightly increasing for A/H1N1pdm (see supplement).
Hence, the fact that this average rose for $t<0$ gives little information for $t>0$, and is due to the conditions by which these trajectories were selected.
This shows that sweep-like trajectories rising steadily from frequency $0$ to $1$ are not common enough to dominate the average trajectory.  \pierre{This result is consistent with\citep{illingworth_components_2012}. They show that sweep-like trajectories are very bad fits to actual trajectories.}

@@

This dynamics is in apparent contradiction with evidence that influenza surface proteins are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
If strong selection was present, we would expect rising amino acid mutations to fix at a distinctively higher frequency than the one at which they are measured.
In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.

Next, we searched for features of mutations that allow prediction of fixation beyond frequency by dividing frequencies into categories that deviate from the diagonal in panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active}.
We first turn to the \emph{Local Branching Index} (LBI), a quantity calculated for each node in a phylogenetic tree that indicates how dense the branching of the tree is around that node.
LBI has previously been successfully used as a predictor of the future population of influenza~\cite{neher_predicting_2014}, and was shown to be a proxy for fitness of leaves or ancestral nodes in mathematical models of evolution.
Here, we define the LBI of a mutation at date $t$ as the average LBI of strains that carry this mutation and that were sampled in the time bin corresponding to $t$.
Panel A of figure~\ref{fig:pfix_stratified} shows fixation probability for HA mutations with LBI in the top or bottom half of the distribution.
Both groups have identical probability of fixation, suggesting that LBI carries very little information on the probability of fixation of a mutation.

@@

In supplementary figure \sref{fig:pfix_v_freq_combijohnsmodels}, we repeat this analysis for several predictors from our recent comparison of predictive models for influenza \citep{huddleston_integrating_2020} including models that use serological data.
While many of these model predict future population composition, the only score that was predictive of fixation is mutations load.

@@

Predicting the trajectory of a mutation requires (i) significant fitness difference between genomes carrying different variants at the site and (ii) a selection pressure that changes slowly over time.
Under such conditions, it is expected that frequency trajectories will show a persistent behavior which would make them predictable for some time.
However, we could find only limited evidence for such persistent behavior in the past 19 years of IAV evolution.
This lead us to conclude that (i) influenza virus evolution is qualitatively different from models of rapidly adapting population (despite clear evidence for frequent positive selection), and (ii) most methods to predict influenza evolution are unable to predict future frequency dynamics even though they work well for predicting the composition of the future population. 
This is illustrated by the case of LBI, which works primarily because it picks strains that represent the future well and not because it is predictive of future dynamics.
% work primarily because they pick strains that represent the future well, not because they predict future dynamics.

The primary focus in this work was the investigation of frequency trajectories of new amino acid mutations.
In the short term, we found that on average the direction of trajectory does not persist for longer than a few months.
Indeed, the average trajectory in figure \ref{fig:trajectories} takes a sharp turn when going from $t<0$ to $t>0$, instead of showing ``inertia''.
This suggests that selective sweeps are not representative of typical trajectories. \pierre{We could cite the Illingworth paper here. They do show that as well, in a different way. The fact many trajectories are not sweeps is expected when there is strong linkage. What cannot be easily explained by linkage only is the complete lack of inertia on average (at least I think?), and the $P_{fix}(f)=f$. }

On a longer timescale, we investigated the probability that a novel mutation observed at frequency $f$ fixes.
In neutral models of evolution this probability equals $f$, while it should be higher or lower than $f$ for mutations with a beneficial or deleterious effect on fitness, respectively.
However, in the case of influenza, this probability differs little from $f$, making current frequency the best predictor for fixation.
In figure~\ref{fig:pfix_stratified}, we split trajectories into groups for which we expected $P_{fix}$ to deviate from $f$.
Many of these splits, such as high/low LBI or epitope/non-epitope positions, did not result in an increased predictability, while others gave limited information on fixation.
Despite the lack of predictability of mutation frequency trajectories, influenza surface proteins show strong signatures of selection \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.

>>>>

We focus on frequency trajectories that are starting at a zero (low) frequency, \emph{i.e.} $f(t=0)=0$.
These represent new amino acid variants which were absent in the population at the time bin when the trajectory started and are currently rising in the population (see Methods).
Such distinction in novel and ancestral variants is necessary to meaningfully interrogate predictability.
Each rising trajectory of a new mutation implies the existence of another decreasing one at the same position, since frequencies of all amino acids at a given position must sum to one.
%If novel variants arise by selection, we expect to see a stronger signal of selection after conditioning on these novel variants.
In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e., the rapid rise and fix.
The sweep of a mutation can be due to its own fitness effect, to the genetic background or to the effect of the seven other segments.
By considering the ensemble of novel variants that are rising in frequency, we effectively average over backgrounds, obtaining a set of mutations that we expect to be beneficial on average.
If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency should thus enrich for mutations that are positively selected and on their way to fixation.
Interference between multiple such advantages mutations will perturb the trajectories \citep{illingworth_components_2012} but we still expect an average \pierre{replaced ``consistent'' by ``average''} tendency to rise in frequency (see below). \pierre{Is this ``see below'' necessary? I think that's what the reader naturally does! }

@@

However, future distributions for $\Delta t >0$ do not seem to follow a pattern compatible with selective sweeps.
The thick black line in Figure~\ref{fig:trajectories}B shows the average frequency of all trajectories.
This average makes a sharp turn at $t=0$ and is essentially flat for $t>0$ in the case of A/H3N2, and slightly increasing for A/H1N1pdm (see supplement).
Hence, the fact that this average rose for $t<0$ gives little information for $t>0$, and is due to the conditions by which these trajectories were selected.
This shows that sweep-like trajectories rising steadily from frequency $0$ to $1$ are not common enough to dominate the average trajectory.

@@

This dynamics is in apparent contradiction with evidence that influenza surface proteins in A/H3N2 are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
If strong selection was present, we would expect rising amino acid mutations to fix at a distinctively higher frequency than the one at which they are measured.
In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.

Next, we searched for features of mutations in A/H3N2 that allow prediction of fixation beyond frequency by dividing frequencies into categories that deviate from the diagonal in panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active}.
We first turn to the \emph{Local Branching Index} (LBI), a quantity calculated for each node in a phylogenetic tree that indicates how dense the branching of the tree is around that node.
LBI has previously been successfully used as a predictor of the future population of influenza~\cite{neher_predicting_2014}, and was shown to be a proxy for fitness of leaves or ancestral nodes in mathematical models of evolution.
Here, we define the LBI of a mutation at date $t$ as the average LBI of strains that carry this mutation and that were sampled in the time bin corresponding to $t$.
Panel A of figure~\ref{fig:pfix_stratified} shows fixation probability for HA mutations with LBI in the top or bottom half of the distribution.
Both groups have identical probability of fixation, suggesting that LBI carries very little information on the probability of fixation of a mutation.

@@

In supplementary figures \sref{fig:pfix_v_freq_indivjohnsmodels} and \sref{fig:pfix_v_freq_combijohnsmodels}, we repeat this analysis for the best predictors from our recent comparison of predictive models for influenza \citep{huddleston_integrating_2020}.
We consider individual predictors based on LBI, mutational load, variation in frequency and serological data, as well as two combinations of these quantities. 
While many of these models predict future population composition, the only score that was predictive of fixation is mutational load.

@@

Predicting the trajectory of a mutation requires (i) significant fitness difference between genomes carrying different variants at the site and (ii) a selection pressure that changes slowly over time.
Under such conditions, it is expected that frequency trajectories will show a persistent behavior which would make them predictable for some time.
However, we could find only limited evidence for such persistent behavior in the past 19 years of IAV evolution.
\pierre{We could change point (i) below to only mention A/H3N2. }
This lead us to conclude that (i) influenza virus evolution is qualitatively different from models of rapidly adapting population (despite clear evidence for frequent positive selection), and (ii) many methods to predict influenza evolution are unable to predict future frequency dynamics even though they work well for predicting the composition of the future population. 
This is illustrated by the case of LBI, which works primarily because it picks strains that represent the future well and not because it is predictive of future dynamics.
% work primarily because they pick strains that represent the future well, not because they predict future dynamics.

The primary focus in this work was the investigation of frequency trajectories of new amino acid mutations.
In the short term, we found that on average the direction of trajectory does not persist for longer than a few months.
Indeed, the average trajectory in figure \ref{fig:trajectories} takes a sharp turn when going from $t<0$ to $t>0$, instead of showing ``inertia''.
This suggests that selective sweeps are not representative of typical trajectories. \pierre{We could cite the Illingworth paper here. They do show that as well, in a different way.}

On a longer timescale, we investigated the probability that a novel mutation observed at frequency $f$ fixes.
In neutral models of evolution this probability equals $f$, while it should be higher or lower than $f$ for mutations with a beneficial or deleterious effect on fitness, respectively.
However, in the case of influenza, this probability differs little from $f$, making current frequency the best predictor for fixation.
In figures~\ref{fig:pfix_stratified} and~\sref{fig:pfix_v_freq_indivjohnsmodels} , we split trajectories into groups for which we expected $P_{fix}$ to deviate from $f$.
Many of these splits, such as high/low LBI or epitope/non-epitope positions, did not result in an increased predictability, while others gave limited information on fixation.
Despite the lack of predictability of mutation frequency trajectories, influenza surface proteins show strong signatures of selection \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
